Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Word Processors

### 2018-08-01 main.tex
Implemented HSC comments

\usepackage[
	backend		= biber,
	style		= authoryear-ibid,
	giveninits	= true, 	uniquename	= init, 	natbib		= true,
	url			= false,
	isbn			= false,
	doi			= false,
	eprint		= false,
	maxbibnames	= 99,
	minbibnames	= 3
]{biblatex}

@@

In general, the translator is drafted to translate ambiguous expressions or automatically find a work-around to disambiguate the expression {\bf [HSC: I don't undestand this sentence. drafted? Please rewrite]}. Only in case there is no way to solve the ambiguity with the defined rules, the translation process stops.

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balancing of parentheses in expressions. An expression such as `$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. But this functionality is not implemented yet. {\bf [HSC: This can be solved by the use of interval macros. e.g., 
\verb|\intcc@{a}{b} = [a,b]|, \verb|\intoc@{a}{b}=(a,b]|, \verb|\intco@{a}{b}=[a,b)|, \verb|\intoo@{a}{b}=(a,b)|,
with a double \verb|@@| notation you could do other notations such as \verb|\intoo@@{a}{b}=]a,b[|]}

The \verb|SequenceTranslator| also handles positions of multiplication symbols. There are a couple of obvious choices to translate multiplication. The most common symbol for multiplications is still the white space or none symbol {\bf [HSC: What is a none symbol?]} at all, as explained previously. Consider the simple expression `$2n\pi$'. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as `$f(x)$'. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.

@@

Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has. In consequence, it also processes the following siblings to translate the arguments. There are some special cases about the next node right after the macro occurs. This node can be
\begin{itemize}
\item an exponent, such as for `\verb|^2|';
\item an optional parameter in square brackets;
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above;
\item `$@$' symbols if none of the above {\bf [HSC: it was originally written ``an @ symbols'' which doesn't make
sense, please rewrite if wrong.]}; or
\item a variable in curly brackets (a \textit{sequence} node) if none of the above.
\end{itemize}

@@

One can access the internal data structure of expressions via the \texttt{ToInert} command. The returned \inertF{} is a nested list representation as a tree equivalent {\bf [HSC: I don't understand the previous sentence. Please rewrite.]}\footnote{A tree equivalent of a \gls*{dag} splitting nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in \Cref{tab:maple-types}.
\begin{table}[h]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums. Internally stored with factors for each summand, i.e., `$x+y$' would be stored as `$x \cdot 1 + y \cdot 1$'.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} data types. See~\parencite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}
The translator uses the \texttt{OpenMaple}~\parencite[\S 14.3]{MAPLE:ProgrammingGuide} \gls*{api} for interacting with \Maple's kernel and get access {\bf [HSC: and get access?? please rewrite.]}{\bf [HSC: and get access?? please rewrite.]}  to the \inertF{} of inputs.

@@

The round trip tests are very successful, but they only detect errors in string representations. However, because of the simplification techniques of fixed points, we are able to at least detect logical errors in one system: \Maple. On the other hand, these tests cannot determine logical errors in the translations between the two systems. Consider we mistakenly defined an \textit{inappropriate} forward and backward translation for the sine function
\begin{eqnarray}
\verb|\sin@{\phi}| & \overset{\langMaple}{\leftrightarrow} & \verb|cos(phi)|,\label{eq:wrong-trans-1}\\
\verb|\cos@{\phi}| & \overset{\langMaple}{\leftrightarrow} & \verb|sin(phi)|.\label{eq:wrong-trans-2}
\end{eqnarray}
In that case the round trip test would not detect any errors and reach a fixed point representation, because the simplification techniques only simplify two representations in the same system but cannot compare the representation in one system to those in the other {\bf [HSC: This sentence is too long and rambles on and is not easy to read Can you break it into two sentences and rewrite them so that it is easier to read?]}.

@@

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls*{cas} often have difficulties testing simple equations {\bf [HSC: Can you give an example of this?]}. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \texttt{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \texttt{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. Because of implementation details, there are some techniques that helps \Maple{} to find possible simplifications. For example we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an equivalent representation using their exponential representations, namely
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions, we are able to improve the simplification process in \Maple. However, the limitations of the \texttt{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ is the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains a $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one should instead use the analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the equivalence checks of \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

The \gls*{dlmf} provides a dataset of \LaTeX{} expressions with semantic macros. We extracted 4087 equations from the \gls*{dlmf} and apply our round-trip and relation tests on them. The translator was able to translate 2405\footnote{All percentages are approximately calculated.} (58.8\%) of the extracted equations without errors. 660 (27.4\%) of the successfully translated expressions were verified by the simplification techniques of \Maple. We apply additional numerical tests for the remaining 1745 equations. For 418 (24\%) cases the numerical tests were valid. More detailed results for numerical and symbolical tests were presented in~\parencite{NumericalTests:Paper} {\bf [HSC: Why is this rendered as (ibid.)? Remove when fixed.]}.

@@

We have also noticed that our testing procedure is able to identify errors in CAS procedures,
namely the \Maple{} {\tt simplify} procedure. The left-hand side of~\parencite[(7.18.4)]{NIST:DLMF} 
{\bf [HSC: Why is this rendered as ibid? Remove when fixed.]} is given by 
\begin{equation*}
\frac{{\mathrm{d}}^{n}}{{\mathrm{d}z}^{n}}\left(e^{z^{2}}\operatorname{erfc}z\right),
\end{equation*}
where $n \in \mathbb{N}_0$, $e$ is the base of the natural logarithm, and $\operatorname{erfc}$ is the
complementary error function. Our translation correctly produces 
\begin{equation*}
\verb|diff((exp((z)^(2))*erfc(z)), [z$(n))]|.
\end{equation*}
However, the \Maple{ 2016} \verb|simplify| function falsely returns $0$ for the translated left-hand side.
Maplesoft has confirmed in a private communication that this is indeed a defect in \Maple{ 2016}.
Furthermore, although the nature of the defect changes, the defect still persists in \Maple{ 2018}
as of the publication of this manuscript.

\section{Conclusion \& Future Work}\label{ch:conc-future-work}
The results and conclusions of the translations of the formula set from the \DLMF{} are diverse. Some test cases have shown us that the set of semantic macros are not perfect yet and need to be improved. This is seen in the example of the macro definition for the Wronskian determinant. However, simply adding more and more macros to improve the set is not the right decision. In that case, we would end up in something similar to \sTeX, which is overloaded and too complicated to use. If the system is no longer handy, it will not be used {\bf [HSC: I don't agree with this paragraph. In fact, I think we've had this discussion already in the past and it was resolved. It needs to be resolved again. There are many examples where enhancing macros is a benefit. Not only that, we've discussed this recently in our publications. This needs to be fixed prior to submission.]}

On the other hand, the translator concept has proven itself by discovering errors in the online \gls*{dlmf} compendia. The test cases have also shown how difficult it is to validate a translated expression and have uncovered the problems of translations between two systems with different sets of supported functions. Our validation techniques also assume the correctness of the simplification and computational algorithms in \gls*{cas}. However, combining those techniques and automatically running translation checks, not only can discover errors in mathematical compendia but can also detect errors in simplifications or computations of the \gls*{cas}.

@@

The translator was designed to be easily extendable. This allows one to implement translations for other \gls*{cas} without much effort.
The biggest weakness of the translator is still the semantic macros {\bf [HSC: I don't understand what this sentence means. I think you need to be more specific here. The way it is written it doesn't make sense.]}. Currently, we are working on mathematical information retrieval techniques which will allow for an extension of the translator to generic \LaTeX{} inputs.

{\bf [HSC: The link in the NIST:DLMF reference is broken (see below), and I don't know how to fix it. By the way, I made some changes the the way that it is referred to in the text, because I did not like that Olver et al was being referred to, since it is a web site. However, it is still not being displayed properly. For instance, the Version number is not being displayed and also the link is broken. Can someone try to fix this? I do not know Biber, so this is probably the reason for this mess.]}

>>>>

\usepackage[
	backend		= biber,
	style		= authoryear,
	giveninits	= true, 	
	uniquename	= init, 	
	natbib		= true,
	url			= false,
	isbn			= false,
	doi			= false,
	eprint		= false,
	maxbibnames	= 99,
	minbibnames	= 3
]{biblatex}

@@

In general, the translator is designed to find a work-around for disambiguating expressions. Only in case there is no way to solve the ambiguity with the defined rules, the translation process stops.

@@

A \verb|SequenceTranslator| translates the \textit{sequence} and \textit{balanced expressions} in the \gls*{pom-pt}. If a node $n$ is a leaf and the represented symbol is an open bracket (parentheses, square brackets and so on) the following nodes are also taken as a \textit{sequence}. Hence, combined with the recursive translation approach, the \verb|SequenceTranslator| also checks balancing of parentheses in expressions. An expression such as `$(a]$' is producing a mismatched parentheses error. On the other hand, this is a problem for interval expressions such as `$[a,b)$'. In the current version, the program cannot distinguish between mismatched parentheses and half-opened, half-closed intervals. Whether an expression is an interval or another expression is difficult to decide and can depend on the context. Also, the parentheses checker could simply be deactivated to allow mismatched parentheses in an expression. Another option is to use of interval macros. e.g., \verb|\intcc@{a}{b} = [a,b]|.

The \verb|SequenceTranslator| also handles positions of multiplication symbols. There are a couple of obvious choices to translate multiplication. The most common symbol for multiplications is still the white space (or no space between the tokens), as explained previously. Consider the simple expression `$2n\pi$'. The \gls*{pom-pt} generates a sequence node with three children, namely $2$, $n$ and $\pi$. This sequence should be interpreted as a multiplication of the three elements. The \verb|SequenceTranslator| checks the types of the current and next nodes in the tree to decide if it should add a multiplication symbol or not. For example, if the current or next node is an operator, a relation symbol or an ellipsis, there will be no multiplication symbol added. However, this approach implies an important property. The translator interprets all sequences of nodes as multiplications as long as it is not defined otherwise. This potentially produces strange effects. Consider an expression such as `$f(x)$'. Translate this to \Maple{} will be \verb|f*(x)|. But we do not consider this translation to be wrong, because there is a semantic macro to represent functions. In this case, the user should use \verb|\f{f}@{x}| instead of \verb|f(x)| to distinguish between $f$ as a function call and $f$ as a symbol.

@@

Algorithm~\ref{alg:macro-translation} explains the \verb|MacroTranslator| without error handling. It has extracted necessary information from the \gls*{pom-pt}, such as how many arguments this function has. In consequence, it also processes the following siblings to translate the arguments. There are some special cases about the next node right after the macro occurs. This node can be
\begin{itemize}
\item an exponent, such as for `\verb|^2|';
\item an optional parameter in square brackets;
\item a parameter in curly brackets (a \textit{sequence} node in the \gls*{pom-pt}) if none of the above;
\item `$@$' symbols if none of the above; or
\item a variable in curly brackets (a \textit{sequence} node) if none of the above.
\end{itemize}

@@

One can access the internal data structure of expressions via the \texttt{ToInert} command, which returns the \inertF. The \inertF{} format is a nested list\footnote{The nested list is a tree representation of a \gls*{dag} that splits nodes with multiple parents into multiple nodes so that each node has only one parent node.} of the internal \gls*{dag} for the given expression. Some of the important types for the nodes are specified in \Cref{tab:maple-types}.
\begin{table}[h]
\centering
\begin{tabular}{cp{10cm}}
	\hline
	Type & Explanation\\
	\hline
	SUM & Sums. Internally stored with factors for each summand, i.e., `$x+y$' would be stored as `$x \cdot 1 + y \cdot 1$'.\\
	PROD & Products.\\
	EXPSEQ & Expression sequence is a kind of list. The arguments of functions are stored in such sequences.\\
	INTPOS & Positive integers.\\
	INTNEG & Negative integers.\\
	COMPLEX & Complex numbers with real and imaginary part.\\
	FLOAT & Float numbers are stored in the scientific notation with integer values for the exponent $n$ and the significand $m$ in $m \cdot 10^n$.\\
	RATIONAL & Rational numbers are fractions stored in integer values for the numerator and positive integers for the denominator.\\
	POWER & Exponentiation with expressions as base and exponent.\\
	FUNCTION & Function invocation with the name, arguments and attributes of the function.\\
	\hline
\end{tabular}
\caption{A subset of important internal \Maple{} data types. See~\parencite{MAPLE:ProgrammingGuide} for a complete list.}
\label{tab:maple-types}
\end{table}
The translator uses the \texttt{OpenMaple}~\parencite[\S 14.3]{MAPLE:ProgrammingGuide} \gls*{api} for interacting with \Maple's kernel implementation.

@@

The round trip tests are very successful, but they only detect errors in string representations. However, because of the simplification techniques of fixed points, we are able to at least detect logical errors in one system: \Maple. On the other hand, these tests cannot determine logical errors in the translations between the two systems. Consider we mistakenly defined an \textit{inappropriate} forward and backward translation for the sine function
\begin{eqnarray}
\verb|\sin@{\phi}| & \overset{\langMaple}{\leftrightarrow} & \verb|cos(phi)|,\label{eq:wrong-trans-1}\\
\verb|\cos@{\phi}| & \overset{\langMaple}{\leftrightarrow} & \verb|sin(phi)|.\label{eq:wrong-trans-2}
\end{eqnarray}
In that case the round trip test would not detect any errors but reaches a fixed point representation.

@@

However, there are two essential problems with this approach. Testing the mathematical equivalence of expressions is hard to solve and \gls*{cas} often have difficulties testing simple equations symbolically. For example, equation~\parencite[(4.35.34)]{NIST:DLMF}
\begin{equation*}
\sinh \left( x+\iunit \cdot y \right) = \sinh x \cos y + \iunit \cosh x \sin y,
\end{equation*}
as a difference of the left- and right-hand side cannot be simplified to zero by default. Furthermore, this approach only checks forward translations because there is no way to check equivalence of expressions in \LaTeX{} automatically (again this could become feasible with our translator). We use \Maple's \texttt{simplify} function to check if the difference of the left-hand side and the right-hand side of the equation is equal to zero. In addition, we use \texttt{simplify} and check if the division of the right-hand side by the left-hand side returns a numerical value or not. This simplification function is the most powerful function to check the equivalence in \Maple. However, there are several cases where the simplification fails. Because of implementation details, there are some techniques that helps \Maple{} to find possible simplifications. For example we can force \Maple{} to convert the formula
\begin{equation}
\sinh{x} + \sin{x}
\end{equation}
to an equivalent representation using their exponential representations, namely
\begin{equation}
\frac{1}{2}\expe^x - \frac{1}{2}\expe^{-x} - \frac{1}{2} \iunit \left( \expe^{\iunit x}-\expe^{-\iunit x} \right).
\end{equation}
With such pre-conversions, we are able to improve the simplification process in \Maple. However, the limitations of the \texttt{simplify} function are still the weakest part of this verification approach. Consider the complex example~\parencite[(12.7.10)]{NIST:DLMF}
\begin{equation}\label{eq:branch-cut-near}
\displaystyle U(0,z) = \sqrt{\frac{z}{2\cpi}} K_{\frac{1}{4}}\left(\frac{1}{4}z^2\right),
\end{equation}
where $U(0,z)$ is the parabolic cylinder function and $K_\nu(z)$ is the modified Bessel function of the second kind. Both functions are well-defined in both systems and we can define a \textit{direct} translation for~\eqref{eq:branch-cut-near}. 
The modified Bessel function of the second kind has its branch cut in \Maple{} and in the \gls*{dlmf} at $z < 0$. However, the argument of $K$ contains a $z^2$. If $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ the value of the right-hand side of~\eqref{eq:branch-cut-near} would be no longer on the principal branch. However, \Maple{} will still compute the principal values independently of the value of $z$. Hence, a translation
\begin{equation}
\verb|\BesselK{\frac{1}{4}}@{\frac{1}{4}z^2}| \overset{\langMaple}{\mapsto} \verb|BesselK(1/4,(1/4)*z^2)|
\end{equation}
is incorrect if $|\ph{z}| \in \left(\frac{\cpi}{2}, \cpi\right)$ and one should instead use the analytic continuation for the right-hand side of (\ref{eq:branch-cut-near}). 
To evaluate such complex cases, the equivalence checks of \gls*{cas} are insufficient. Therefore we implement numerical tests as an additional step.

@@

The \gls*{dlmf} provides a dataset of \LaTeX{} expressions with semantic macros. We extracted 4087 equations from the \gls*{dlmf} and apply our round-trip and relation tests on them. The translator was able to translate 2405\footnote{All percentages are approximately calculated.} (58.8\%) of the extracted equations without errors. 660 (27.4\%) of the successfully translated expressions were verified by the simplification techniques of \Maple. We apply additional numerical tests for the remaining 1745 equations. For 418 (24\%) cases the numerical tests were valid. More detailed results for numerical and symbolical tests were presented in~\parencite{NumericalTests:Paper}.

@@

We have also noticed that our testing procedure is able to identify errors in CAS procedures,
namely the \Maple{} {\tt simplify} procedure. The left-hand side of~\parencite[(7.18.4)]{NIST:DLMF} is given by 
\begin{equation*}
\frac{{\mathrm{d}}^{n}}{{\mathrm{d}z}^{n}}\left(e^{z^{2}}\operatorname{erfc}z\right),
\end{equation*}
where $n \in \mathbb{N}_0$, $e$ is the base of the natural logarithm, and $\operatorname{erfc}$ is the
complementary error function. Our translation correctly produces 
\begin{equation*}
\verb|diff((exp((z)^(2))*erfc(z)), [z$(n))]|.
\end{equation*}
However, the \Maple{ 2016} \verb|simplify| function falsely returns $0$ for the translated left-hand side.
Maplesoft has confirmed in a private communication that this is indeed a defect in \Maple{ 2016}.
Furthermore, although the nature of the defect changes, the defect still persists in \Maple{ 2018}
as of the publication of this manuscript.

\section{Conclusion \& Future Work}\label{ch:conc-future-work}
%The results and conclusions of the translations of the formula set from the \DLMF{} are diverse. Some test cases have shown us that the set of semantic macros are not perfect yet and need to be improved. This is seen in the example of the macro definition for the Wronskian determinant. However, simply adding more and more macros to improve the set is not the right decision. In that case, we would end up in something similar to \sTeX, which is overloaded and too complicated to use. If the system is no longer handy, it will not be used {\bf [HSC: I don't agree with this paragraph at all. In fact, I think we've had this discussion already in the past and it was resolved. It needs to be resolved again. There are many examples where enhancing macros is a benefit. Not only that, we've discussed this recently in our publications. This needs to be fixed prior to submission. I vote for removing this paragraph altogether until we can resolve this issue.]} {\bf [AGP: Since the paper in it's current state is quite long and the section works well without this paragraph, I just commented it out and avoid the discussion here...]}

The translator concept has proven itself by discovering errors in the online \gls*{dlmf} compendia. The test cases have also shown how difficult it is to validate a translated expression and have uncovered the problems of translations between two systems with different sets of supported functions. Our validation techniques also assume the correctness of the simplification and computational algorithms in \gls*{cas}. However, combining those techniques and automatically running translation checks, not only can discover errors in mathematical compendia but can also detect errors in simplifications or computations of the \gls*{cas}.

@@

The translator was designed to be easily extendable. This allows one to implement translations for other \gls*{cas} without much effort.
An extensive weakness is the dependency on the special macros from the \gls*{dlmf}. The translator is not able to translate functions without using these macros. Currently, we are working on mathematical information retrieval techniques which will allow for an extension of the translator to generic \LaTeX{} inputs.
